phRAIDER: Pattern-Hunter based Rapid Ad Initio \\ Identification of Elemenrary Repeats

### 2015-09-17 manuscript/JCB/manuscript.tex
Updating manuscript

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

% Added by karro for commenting -- should be removed before submission.
\usepackage{color}
\newcommand{\red}[1]{{\color{red}#1}}
\newcommand{\blue}[1]{{\color{blue}#1}}
\newcommand{\cyan}[1]{{\color{cyan}#1}}
\newcommand{\magenta}[1]{{\color{magenta}#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Fast de novo transposable element annotation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},
   noteref={n1},
   email={schaefce@miamiOH.edu}
]{\inits{SE}\fnm{Carly E} \snm{Schaeffer}}
\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   noteref={n1},                        % id's of article notes, if any
   email={figuernd@miamiOH.edu}   % email address
]{\inits{ND}\fnm{Nathaniel D} \snm{Figueroa}}
\author[
   addressref={aff1,aff2},
   email={liux17@miamiOH.edu}
]{\inits{X}\fnm{Xiaolin} \snm{Liu}}
\author[
  addressref={aff1,aff2,aff3,aff4},
  corref={aff1},
  noteref={n2},
  email={karroje@miamiOH.edu}
]{\inits{JE}\fnm{John E} \snm{Karro}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Department of Computer Science and Software Engineering}, % university, etc
  %\street{Miami University},                     %
  %\postcode{}                                % post or zip code
  %\city{Oxford},                              % city
  %\cny{USA}                                    % country
}
\address[id=aff2]{%                           % unique id
  \orgname{Center for  Molecular and Structural Biology},
  %\street{Miami University},                     %
  %\postcode{}                                % post or zip code
  %\city{Oxford},                              % city
  %\cny{USA}                                    % country
}
\address[id=aff3]{%                           % unique id
  \orgname{Department of Statistics}, % university, etc
  %\street{Miami University},                     %
  %\postcode{}                                % post or zip code
  %\city{Oxford},                              % city
  %\cny{USA}                                    % country
}
\address[id=aff4]{%                           % unique id
  \orgname{Department of Microbiology}, % university, etc
  \street{Miami University},                     %
  %\postcode{}                                % post or zip code
  \city{Oxford},                              % city
  \cny{USA}                                    % country
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
  \note[id=n1]{Equal contributor} % note, connected to author
  \note[id=n2]{Corresponding Author}
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
\parttitle{Background} The problem of {\it de novo} identification of
transposable elements (the discovery and annotation of transposable
elements without the use of a pre-complied profile) has been addressed
by a number of tools -- all of which are either based on
computationally complex algorithms that cannot scale to whole-genome
use, or whose sensitivity suffers significantly from the presence of
sequence variation.  Here we present phRAIDER (Pattern Hunter based
Rapid Ab Initio Detection of Elementary Repeats), a tool that can
quickly identify and mask tranposable elements in a a newly sequenced
genome and is robust to the sequence variation present in real data.

\parttitle{Results} \red{To be written later.}

\parttitle{Conclusions} \red{To be written later.}

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{transposable elements}
\kwd{elementary repeats}
\kwd{pattern hunter}
\kwd{genomic masking}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newtheorem{definition}{Definition}
\newtheorem{observation}{Observation}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
\section*{Background}
Transposable Elements (TEs) are genomic sequences thathad the capacity
to insert copies of themselves into other genomic locations, resulting
in homologous families of sequences spread across the genome.  Present
in almost every higher order genome (covering as much as 45\% of the
human genome and 90\% of the maize genome
\cite{Venter:2001p92,SanMiguel:1996wa}), TEs have proved an important
source of data in numerous studies of genomic structure (e.g.
\cite{Arndt:2005p279,Karro:2008p123,Mugal:2009p581,Hardison:2003p97}).
But given their prevalence, it is important for those studying other
aspect the genome to have TEs masked out -- their bases replaced by N
to allow for easy identification and filtering.  Failure to filter can
reek havoc with genomic analysis tools.  For example, unfiltered TEs can trigger
huge numbers of false positives in automated gene finding tool
\cite{Jiang:2013jt}, as well as inflate tool runtime.

@@

Within mammalian species we can largely rely on homology relationships
to port libraries across species.  This does not hold
so well in plants: in many cases TE composition of a given plant
organism is species-specific.  For example, a rice-based TE library
will only identify 25\% of the TEs in the maize genome
\cite{Jiang:2013jt}.

To solve this problem we turn to {\it de novo} TE identification
tools, identifying TEs using only the genome sequence information.  A
number of such tools are discussed in the literature.  RECON uses
WU-BLAST and PILER using LASTZ to compute self-alignments
\cite{Bao:2002,Edgar:2005p2365,Lopez:2003td,Harris:2007uf}.  RECON
show good sensitivity but is computationally intensive and infeasible
for use on whole genomes (requiring 60 hours for 18Mb rice genome in a
2013 study), while PILER achieves a good runtime with very low
sensitivity .  ReAS and RepeatScout \cite{Price:2005p1247} are based
on $k$-mer searches, with the earlier showing less sensitivity than
RECON \cite{Li:2005he, Price:2005p1247,Jiang:2013jt}.  RepeatGluer
\cite{Pevzner:2004p3157,Zhi:2006p3199} is based on a variation of
DeBrujin graphs, which allows for a decomposition of TE families into
domains, but is very, very computationally expensive.  In Saha {\it et
  al.} the authors perform an extensive comparison of the tools, and
conclude that RepeatScout is the best tool overall for assembled
genomes, while ReAS the best when dealing with unassembled sequence
fragments \cite{Saha:2008dm}.

\subsection*{Elementary Repeats}
Zheng and Lonardi approached the {\it de novo} identification problem using {\it elementary repeats}
\cite{Zheng:2005bl}.
Similar to the RepeatGluer domains, elementary repeats are
decompositions of TEs into basic building blocks.  Identification of
these building blocks are sufficient for the purpose
of masking, and can be assembled into Transposable Elements for those
interested in TEs themselves.

While it is notoriously difficult to mathematically model transposable
elements \cite{Bao:2002}, elementary repeats are more conducive to a
formal description.  For a given genome, a nucleotide sequence $r$ is
an elementary repeat if: (1) It is of at least length $l$ (the length
requirement); (2) there are at least $f$ copies of $r$ appear (the
frequency requirement) (3) there is no proper substring of $r$ of
length $\leq l$ that appears in the genome independently of $r$ (the
minimality requirement); (4) $r$ is a maximal string w.r.t (1-3) (the
maximality requirement).  Having proposed this definition, Zheng and
Lonardi developed an identification algorithm that had a runtime
quadratic in the query sequence size \cite{Zheng:2005bl}. This was
refined to linear time by He and also by Huo {\it et al.}
\cite{He:2006gpa,Huo:2009hoa} based on variations of suffix tree
approaches, but these appraoches are limited in their ability to handle sequence
variation.  As we are looking a genome size inputs, and TEs
transposable elements naturally suffer from copy mistakes and
accumulate instance-specific base substitutions over time, this is a
significant limitation.

\subsection*{RAIDER}
\label{RAIDERSec}
It was with the objective of creating a linear time identification
algorithm that could handle variation through use of
PatternHunter-like spaced seeds that we developed the prototype RIADER
\cite{Li:2004wl}.  A rough implementation was first presented in
Figueroa {\it et al.}, with more details in the Figueroa masters
thesis \cite{Figueroa:2014uk,Figueroa:2013cz}.  RAIDER was built along
an alternate, but equivalent definition of elementary repeats based on
$l$-mers (sequences of length $l$).  Specifically, it was observed
that the minimality condition could be rewritten as: There is no
$l$-mer contained within elementary repeat $r$ that appears in the
genome more times than $r$.  From there we make four core observations
that form the basis for the RAIDER algorithm: (1) An $l$-mer cannot
belong to two different elementary repeats; (2) Any $l$-mer in the
genome that occurs $f$ or more times is either an elementary repeat or
belongs to one; (3) any two $l$-mers belonging to an elementary repeat
must appear the same number of times in the genome; (4) If two
sequences in the genome that are {\it maximally identical} (that is,
cannot be extended in either direction and still be identical), these
sequences cannot belong to a larger elementary repeat.  (By ``belong''
we mean ``is a substring of'' -- a definition we will be generalizing
shortly.)  For discussion and proof, see the Figueroa Thesis
\cite{Figueroa:2013cz}.

Based on these observations, we discover we can find all elementary
repeats in a single scan of the genome.  Specifically, as we scan from
left to right, we track $l$-mer occurrences and identify multiple
copies of the same $l$-mer.  When we find the same sequence of $l$-mers
occurring multiple times in a row, we can mark it as a tentative family,
then break it down later if we discover violations of the minimality
condition.  The algorithm is summarized in Figueroa {\it et al.}, and
discussed in detail in the Figueroa Thesis \cite{Figueroa:2014uk,Figueroa:2013cz}.

Results of the preliminary implementation were promising. On human
chromosome 22 we saw a $12\times$ speedup over RepeatScout to RAIDER
(2344 seconds to 192 seconds), while coverage of the RepBase
\cite{Jurka:2005bl} ancestral sequence improved (77\% to 84\%), while
on mouse chromosome 19 we saw the same speedup with a significant drop
in coverage (53\% to 30\%).  On the full human genome RAIDER ran in
$6.3$ hours, while RepeatScout was unable to complete it run.  For
details, see Figueroa {\it et al.} \cite{Figueroa:2014uk}.

\subsection*{Spaced Seeds}

PatternHunter, a very successful augmentation to BLAST
\cite{Li:2004wl,Altschul:1997p843}, is based on the notion of {\it
  spaced seeds}: improving the sensitivity of string matching based
algorthms by allowing wild-cards in the match.  That is, instead of
requiring two strings matching in 12 consecutive characters, we might
instead require two six-character exact matches seperated by one base
which may or may not match (represented by the {\it seed pattern}
11111101111111), or perhaps three consecutive four-character exact
matches sperated by two bases each (1111001111001111). It has been
demonstrated that certain seed patterns can indunce significant
imporvements in BLAST sensitivity with out time penalty, though what
makes a good pattern is not well understood.

RAIDER was designed with the intent of employing the spaced seed
strategy, but this was only implemented heuristically for the Figuera
{\it et al.} paper \cite{Figueroa:2013cz} -- serving primarily as a
proof-of-concept.  Since its release we have develoed a formal model
of elementary repeats that incorporated spaced seeds, and from that
developed phRAIDER (PatternHunter-based RAIDER).  phRAIDER is a fast
tool for the identification and making of transposible elements in
both assembled and unassembled genomes aht outpreforms RptScout and
other establihsed tools. Code is free available under the Gnu GPL
lisence (v. 3) and  may be obtained [NEED WEB ADDRESS].

In the following we will present a new theoretical model for {\it
  seeded elementary repeats}, followed by an algorithm for identifying
them.  In the results section we will show the result of using that
algorithm for masking transp

In our Methods section
we present the new model that allows us to extend RAIDER to correctly
use spaced seeds, and briefly outline the algorithm (with more details
provided in the supplementary materials), with a analysis of phRAIDER
performance in our Results.

\section*{Model}

Our goal is to redefine the concept of transposbile elements to
accomodate a spaced seed strategy.  In the next section we will describe our
identification algorithm, and then quantified its success in 
masking transposable elements. But we will start here with a brief
outline of our theoretical model (with a more detailed description in
the appendix).

We first need to redfine out terminimology regarding elementary
repeats.  Under the Z\&L definition, all instances of one elementary
repeat have the exact same sequence, and hence can be descrbied by a
single string.  As we are allowing for variation, we will describe out
instance set with a {\it sequence descriptor} consisting of bases
letters and the wild-card character * (e.g. AAC*GG would describe a
set of sequences with starting with an AAC, ending with a GG, and
having any character in between).  Given a binary string $s$
representing a spaced seed, we say a sequence descriptor $r$ is
{\it consistent} with $s$ if we can align $s$ to a substring of $r$ such
that every $*$ in $r$ aligns with a $0$ character in $s$. (Hence
$ACG**T*A$ is consistent with the seed $11001$, but not $11011$.)

Given a sequence descriptor $r$, we can {\it decompose} $r$ with
respect to $s$ by taking every length $|s|$ substring of $r$ that is consistant
with $s$, replacing all letters of $r$ that match to a 0 in $s$ with a
$*$, and creating a set from the results.  (Hence for $s = 11011$ and
$r=AA*CCGTT$, the decomposition would be $\{AA*CC, CC*TT\}$.)  We say
$s$ {\it covers} $r$ is every base in $r$ is contained in at least on
string in the decompositio.  (Hence the previous $s$ does not cover $AAA*CC$, as
the first base in not in any of the strings of the decomposition.)

We can now modif the previous definition of elementary repeats as
follows:
\begin{definition} Given a genomic sequence $G$, an integer $f$, and a
  spaced seed $s$, a sequence descriptor $r$ describes an elementary
  repeat if it meets the four (moified) requirements of an elementary repeat:
\begin{itemize}
\item \underline{Structure requriement}: $s$ covers $r$.
\item \underline{Frequency requirement}: There are at least $f$
  substrings of $G$ that match $r$.
\item \underline{Minimality requirement:} For every string $t$ in the
  decomposition of $r$ w.r.t $s$, the number of occurences of $t$ in
  $G$ is equal to the number of occurneces of $r$ in the genome.
\item \underline{Maximality requirement:} There is no sequence
  descriptor $r'$ of $r$ that contains $r$ as a proper substring and
  satisfies conditions 1-1.
\end{itemize}
\end{definition}

\begin{theorem} When the seed $s$ has no 0 characters, this
    definition of elementary repeats is equivilent to the Z\&L
    definition.
\end{theorem}

\subsection*{phRAIDER Algorithm}

\red{
TO DO: Give a pseudocode description of the algoirthm and some minimal
discussion.
}

\subsection*{Analysis}

\red{
TO DO: Describe pra analysis and quality metrics.
}

\section*{Results}

\red{
Points to be made:
\begin{enumerate}
\item Show: RAIDER2 has better quality masking results then RS.
\item Show: RAIDER2 is better than RAIDER.
\item Show: Seeds matter.
\item Show: Runtime / memory usage
\end{enumerate}

Possible tables / plots:
\begin{itemize}
\item Table showing results on different chromosomes (pra, runtime, memory).
\item Plot of runtime against genome size (simulated data to control size)?
\item Plot of memory usage (simlated data to control size)?
\item We might want to compare the ability to detect transposable
  elements againsts other types of repeats.  (Should be easy to do
  using simulator.)
\end{itemize}
}

\section*{Conclusions}

\red {
Hopefully will beconcluding that RAIDER is a better tool than
RepeatScout.  Posible discussion of tool features.
}

\section*{Appendix}
In order to allow the use of a spaced seed strategy, we will need to
define a new model of elementary repeats that will accommodate it.  In
the following we will first outline our theoretical framework
(described in detail in the Supplementary Materials), and then review
the phRAIDER algorithm.

\subsection*{Seeded Elementary Repeats}
In order to present our new model of elementary repeats, we first need
to define a few terms:
\begin{enumerate}
  \item A spaced seed is a binary string that starts and ends with 1
    symbols (as defined in Li {\it et al.} \cite{Li:2004wl}).
  \item A {\it sequence descriptor} is a DNA string that allows *
    symbols (indicating that a position where base content does not
    matter).  A string matches a sequence descriptor of equal length if
    it has the same base in all non-* positions.
  \item We say the frequency of $r$ in a genomic sequence $G$ (denoted
    $\nu_G(r)$) is the number of subsequences in $G$ that match $r$.
  \item A spaced seed $s$ {\it hits} a substring of sequence
    descriptor $r$ if that substring is the same length of $s$ and,
    when aligned, every $*$ in the substring matches a 0 in the
    string.  (Example: if $s=11011$ and $r=AAA*TTTTT$, then $s$
    hits both $AA*CC$ and $TTTTT$, but not $A*TTT$.)
  \item The {\it decomposition} of $r$ (with respect to seed $s$) is
    the set of all substrings of $r$ that are hit by $s$.  For the
    {\it generalized decomposition}, we take each member of the
    decomposition and replace by $*$ any base that aligned to a 0 in
    the seed.  (Example: if $s=11011$ and $r=AAATTT$, then $AAATT$ is
    in the decomposition, so $AA*TT$ is in the generalized
    decomposition.)
  \item We say a string $s$ {\it covers} a sequence descriptor $r$ if
  every base in $r$ is contained within at least one substring from
  the decomposition of $r$.
\end{enumerate}

\begin{definition}
  \label{REDDef}
  Given a fixed genomic sequence $G$, a integer $f$, and spaced seed $s$, a sequence
  descriptor $r$ is a {\it repeat element descriptor} if:
  \begin{enumerate}
  \item $s$ covers $r$.  (Minimum length requirement.)
  \item $\nu_G(r) \geq f$.  (Frequency requirement.)
  \item For every string $w$ in the generalized decomposition of $r$
    (w.r.t. $s$), $\nu_G(w) = \nu_G(r)$.  (Minimality requirement.)
  \item There is no sequence descriptor $r'$ such that (1) $r$ is
    a substring of $r'$, (2) $\nu_G(r') = \nu_G(r)$, and (3) $r'$
    satisfies conditions 1-3.  (Maximality requirement.)
  \end{enumerate}
\end{definition}

We will refer to the set of sequences in the genome that match the
repeat family descriptor to be the repeat element family of the
descriptor.

\begin{theorem}
Definition~\ref{REDDef} is equivalent to the Z\&L
definition of elementary repeats when $s$ contains no 0 symbols.
\end{theorem}
In the Supplementary materials we prove our definition equivalent to
the Figueroa definition (Definition 5 from the thesis) -- which is
proved to be equivalent to the Z\&L definition in that work \cite{Figueroa:2014uk}.

It is instructive to look at a specific example.  Consider the toy
seed $11011$, let $f=2$, and assume the genome contains a unique
instance of $g_1 = AAAAA{\mathbf C}TTTTT$ and a unique instance of
$g_2 = AAAAA{\mathbf G}TTTTT$.  Consider the sequence descriptor $r_1
= AAAAA*TTTTT$.  A simple inspection shows us that conditions (1)-(3)
of Definition~\ref{REDDef} hold, so if we assume the surrounding
sequence is such that (4) holds as well, this is an elementary repeat
descriptor for this family of sequences.  (Note $AA*AA*TT*TT$ also qualifies -- the
elementary repeat descriptor is not necessarily unique.)  But if we add a second
instance of $g_3 = AAAAAGTTTTT$, something happens that was not
possible in the unseeded version.  While $r_1$ continues to meet the definition, the $g_2/g_3$
subsequence $AAAAGTTTT$ also induces an elementary repeat family
$r_2 = AAAA*TTTT$. (Note that we cannot extend this by, for example,
one base to the left, as the $AAAAA$ substring would cause a violation
of condition (3)).  So $r_2$ matches strings in $G$ that are
substrings of those that $r_1$ matches -- a condition not possible
with the seedless elementary repeats.  This also then violates one of
our core observation on which RAIDER is broken: we now have an $l$-mer
($AAAA*$) that belongs to multiple descriptors.

A second complicating factor is that our repeat element descriptor is
not a sequence of consecutive $l$-mers. In the seedless version, an
elementary repeat of length $n$ can be viewed as a sequence of $n-l+1$
$l$-mers, each overlapping the last by exactly $l-1$ bases. In this
definition we use members of the generalize decomposition, and find
they are not necessarily offset by just one base: given the $g_1$ and
$g_2$ example above, we see that the $r_1 = AAAAA*TTTTT$ descriptor is
composed of $AA*AA$, $AA*TT$, and $TT*TT$ -- each offset from the last
by 3.  As RAIDER assumed that it need a look-back of only one, this
too breaks the algorithm.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
    Text for this section \ldots

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

%% \section*{Figures}
%%   \begin{figure}[h!]
%%   \caption{\csentence{Sample figure title.}
%%       A short description of the figure content
%%       should go here.}
%%       \end{figure}

%% \begin{figure}[h!]
%%   \caption{\csentence{Sample figure title.}
%%       Figure legend text.}
%%       \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
%% \section*{Tables}
%% \begin{table}[h!]
%% \caption{Sample table title. This is where the description of the table should go.}
%%       \begin{tabular}{cccc}
%%         \hline
%%            & B1  &B2   & B3\\ \hline
%%         A1 & 0.1 & 0.2 & 0.3\\
%%         A2 & ... & ..  & .\\
%%         A3 & ..  & .   & .\\ \hline
%%       \end{tabular}
%% \end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \section*{Additional Files}
%%   \subsection*{Additional file 1 --- Sample additional file title}
%%     Additional file descriptions text (including details of how to
%%     view the file, if it is in a non-standard format or the file extension).  This might
%%     refer to a multi-page table or a figure.

%%   \subsection*{Additional file 2 --- Sample additional file title}
%%     Additional file descriptions text.

\end{backmatter}
\end{document}

%%  LocalWords:  de addressref aff neteref Schaeffer noteref Xiaolin
%%  LocalWords:  Liu corref JE Karro Initio phRAIDER genomic TEs Ns
%%  LocalWords:  JIANG RepeatMasker nHMMER BLASTing homology ALU ReAS
%%  LocalWords:  LINEs Jiang PILER RepeatScout mer RepeatGluer Saha
%%  LocalWords:  infeasible DeBrujin et al Zheng Lonardi pre Huo mers
%%  LocalWords:  PatternHunter decompositions subsequence minimality
%%  LocalWords:  maximality RIADER ers substring AACACA AAGACA TACACA
%%  LocalWords:  TTAACTCA TTTTT TTT substrings AAATTT AAATT TT AAAAA
%%  LocalWords:  FIGUREOA AAAAAGTTTTT AAAAGTTTT AAAA TTTT AAAAT novo
%%  LocalWords:  Acknowledgements transposable tranposable runtime
%%  LocalWords:  unassembled RepBase AAGATA

>>>>

\documentclass[10pt]{article}
\usepackage{longtable}
\usepackage{ulem}
\usepackage{authblk}
\usepackage{algorithmic}
\def\includegraphic{}
\def\includegraphics{}
\pagestyle{myheadings}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1ex plus 0.5ex minus 0.2ex}

\usepackage[paper=letterpaper,margin=1in]{geometry}

\title{phRAIDER: Pattern-Hunter based Rapid Ad Initio \\ Identification of Elemenrary Repeats}
\author[1]{Carly E. Schaeffer}
\author[1]{Nathniel D. Figueroa}
\author[2]{Xiaolin Liu}
\author[1,2,3,4]{\\John E. Karro\thanks{Corresponding Author}}
\affil[1]{Department of Computer Science and Software Engineering}
\affil[2]{Cell, Molecular, and Structural Bology}
\affil[3]{Department of Microbiology}
\affil[4]{Department of Statisticcs, Miami University, Oxford, Ohio (USA)}
\renewcommand\Affilfont{\itshape\small}

\date{}

\begin{document}
\maketitle
\abstract{This is the abstract.}

\section*{Introduction}

Transposable Elements (TEs) are genomic sequences thathad the capacity
to insert copies of themselves into other genomic locations, resulting
in homologous families of sequences spread across the genome.  Present
in almost every higher order genome (covering as much as 45\% of the
human genome and 90\% of the maize genome
\cite{Venter:2001p92,SanMiguel:1996wa}), TEs have proved an important
source of data in numerous studies of genomic structure (e.g.
\cite{Arndt:2005p279,Karro:2008p123,Mugal:2009p581,Hardison:2003p97}).
But given their prevalence, it is important for those studying other
aspect the genome to have TEs masked out -- their bases replaced by N
to allow for easy identification and filtering.  Failure to filter can
reek havoc with genomic analysis tools.  For example, unfiltered TEs can trigger
huge numbers of false positives in automated gene finding tool
\cite{Jiang:2013jt}, as well as inflate tool runtime.

@@

\begin{algorithmic}
\If {$i\geq maxval$}
    \State $i\gets 0$
\Else
    \If {$i+k\leq maxval$}
        \State $i\gets i+k$
    \EndIf
\EndIf
\end{algorithmic}

\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc}      % Bibliography file (usually '*.bib' )

\end{document}
